### 2023-05-03 Manuscript/Draft1.Rmd
manuscript update

@@

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values. This can be further examined by viewing yearly estimate/model ratio plots, where we divide the yearly abundance estimates by the true model value and plot them on the log scale. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes. A representative example of this trend is shown in Figure \@ref(fig:RatioTrend).

Analysis of individual yearly model/estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, 33\% of stratified mean estimates were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38. Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16. The breakdown for individual species followed a similar pattern.

When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1. Specifically, when including all strata, the stratified mean produced an average yearly ration of 0.87 with a standard deviation of 0.12, while adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09.  However, when failing to sample the entire domain, adding covariates can decrease the accuracy of VAST estimates. With a reduced number of strata, the stratified mean resulted in our worst ratio estimate with a mean value of 0.59 and standard deviation of 0.10, and adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard devnation of 0.33 to a mean of 0.78 and standard deviation of 0.19.

<font size="5">*Differences Between VAST Settings*</font>

@@

As see in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with $77/80\approx 96\%$ of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covarite response function (linear vs normally distributed etc). 
<!-- **For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information? Which covariate is best? Good question for future research** -->

As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \@ref(fig:strata-plot) shows the strata excluded from sampling and Figure \@ref(fig:CovResponse) depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \@ref(fig:CovResponse), Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

@@

```{r RatioTrend, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of typical ratio trend as shown on a log scale.",out.width="90%"}
# knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
# knitr::include_graphics(c("Images/covariate_est.png"))

knitr::include_graphics(c("Images/ratio_trend.png"))

>>>>

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values. This can be further examined by examining yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can estimate yearly abundance values as low as 0 at times and exhibit large yearly changes. A representative example of this trend for each species is shown on the log scale in Figure \@ref(fig:RatioTrend).

Analysis of individual yearly model:estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38. Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of 0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16. The breakdown for individual species followed a similar pattern.

When the entire domain is sampled, the stratified mean produced an average yearly ratio of 0.87 with a standard deviation of 0.12. Adding covariates in these scenarios brings the VAST estimate ratio closer to 1. Specifically, when all strata are sampled, adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09. Failing to sample the entire domain predictably decreases each individual yearly estimate as the entire population is not being accounted for in the sampling process, which in tern decreases the corresponding estimate ratio. For example, failing to sample the entire domain decreased the estimate ratio for the stratified mean from 0.87 to 0.59 (standard deviation of 0.10). With a reduced number of strata, adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard deviation of 0.33 to a mean of 0.78 and standard deviation of 0.19. As discussed later, when we fail to sample the entire domain in VAST models adding covariates sometimes decreased the accuracy of VAST estimates, typically in the form of additional yearly overestimation.

@@

<font size="5">*Differences Between VAST Settings (mostly notes, not fully formed)*</font>

@@

As see in Table **cite VAST setting table**, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with $77/80\approx 96\%$ of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate. **count how many w/ cov better than w/ out cov?** There is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covariate response function (linear vs normally distributed etc). 
<!-- **For discussion: clearly our perfect covariate information made a big difference for yellowtail flounder (and others), but the question is could we achieve similar results without "perfect" information? Which covariate is best? Good question for future research** -->

As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates. However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \@ref(fig:strata-plot) shows the strata excluded from sampling and Figure \@ref(fig:CovResponse) depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \@ref(fig:CovResponse), Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

@@

```{r RatioTrend, echo=FALSE,out.width="105%",fig.show='hold', fig.align='center', fig.cap="Representative example of typical ratio trend for each species, as shown on a log scale.",out.width="90%"}
# knitr::include_graphics(c("Images/covariate_good.png","Images/covariate_bad.png"))
# knitr::include_graphics(c("Images/covariate_est.png"))
#Yellowtail estimates are for the constant population, constant temperature, all strata used in estimates.
#Cod are Decpop_ConTemp_allStrata_
#Haddock are IncPop_IncTemp_all strata_

knitr::include_graphics(c("Images/ratio_trend_YTF.PNG"))
knitr::include_graphics(c("Images/ratio_trend_cod.PNG"))
knitr::include_graphics(c("Images/ratio_trend_had.PNG"))
```

### 2023-05-03 Manuscript/Draft1.tex
manuscript update

@@

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values. This can be further examined by viewing yearly estimate/model ratio plots, where we divide the yearly abundance estimates by the true model value and plot them on the log scale. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes. A representative example of this trend is shown in Figure \ref{fig:RatioTrend}.

Analysis of individual yearly model/estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, 33\% of stratified mean estimates were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38. Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16. The breakdown for individual species followed a similar pattern.

When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1. Specifically, when including all strata, the stratified mean produced an average yearly ration of 0.87 with a standard deviation of 0.12, while adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09. However, when failing to sample the entire domain, adding covariates can decrease the accuracy of VAST estimates. With a reduced number of strata, the stratified mean resulted in our worst ratio estimate with a mean value of 0.59 and standard deviation of 0.10, and adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard devnation of 0.33 to a mean of 0.78 and standard deviation of 0.19.

\emph{Differences Between VAST Settings}

@@

As see in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. While using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates with \(77/80\approx 96\%\) of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate, there is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covarite response function (linear vs normally distributed etc).

As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates, when a reduced number of strata are sampled adding covariate information leads to a decrease in performance for fall estimates of Haddock and Cod. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \ref{fig:strata-plot} shows the strata excluded from sampling and Figure \ref{fig:CovResponse} depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \ref{fig:CovResponse}, Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

@@

{\centering \includegraphics[width=0.9\linewidth]{Images/ratio_trend}

@@

\caption{Representative example of typical ratio trend as shown on a log scale.}\label{fig:RatioTrend}
\end{figure}

>>>>

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, below the true model values. This can be further examined by examining yearly estimate:model ratio values, where we divide the yearly abundance estimates by the true model value. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can estimate yearly abundance values as low as 0 at times and exhibit large yearly changes. A representative example of this trend for each species is shown on the log scale in Figure \ref{fig:RatioTrend}.

Analysis of individual yearly model:estimate ratios when all strata are included revealed that 73\% of all yearly VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, just 33\% of stratified mean estimate ratios were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios for VAST with spring VAST ratios producing a mean value of 1.08 with a standard deviation of 0.12 while fall VAST ratios were larger with a mean of 1.50 and standard deviation of 0.38. Stratified mean ratios were more consistent between seasons with spring ratios resulting in a mean value of 0.91 with a standard deviation of 0.10, and fall values of providing a mean ratio of 0.84 with a standard deviation of 0.16. The breakdown for individual species followed a similar pattern.

When the entire domain is sampled, the stratified mean produced an average yearly ratio of 0.87 with a standard deviation of 0.12. Adding covariates in these scenarios brings the VAST estimate ratio closer to 1. Specifically, when all strata are sampled, adding covariates improved the VAST estimate ratio from a mean of 1.45 and standard deviation of 0.17 to a mean of 1.13 and standard deviation of 0.09. Failing to sample the entire domain predictably decreases each individual yearly estimate as the entire population is not being accounted for in the sampling process, which in tern decreases the corresponding estimate ratio. For example, failing to sample the entire domain decreased the estimate ratio for the stratified mean from 0.87 to 0.59 (standard deviation of 0.10). With a reduced number of strata, adding covariates to our VAST models decreased the average ratio results from a mean of 1.04 and standard deviation of 0.33 to a mean of 0.78 and standard deviation of 0.19. As discussed later, when we fail to sample the entire domain in VAST models adding covariates sometimes decreased the accuracy of VAST estimates, typically in the form of additional yearly overestimation.

@@

\emph{Differences Between VAST Settings (mostly notes, not fully formed)}

@@

As see in Table \textbf{cite VAST setting table}, VAST models that included covariates used a linear combination of second degree polynomials for habitat and temperature to approximate species-specific covariate responses. Using the exact habitat and temperature covariate information from our population models typically resulted in improved estimates, with \(77/80\approx 96\%\) of VAST models with covariates providing lower absolute error than the corresponding stratified mean estimate. \textbf{count how many w/ cov better than w/ out cov?} There is more to explore with respect to the impact of covariates. For example, by including just one of the covariates individually, one could test which had a larger impact on abundance estimates. Adding noise to our covariate input would test how robust the model-based estimates are to uncertainty in the covariate information. One could also test the impact of assuming the wrong covariate response function (linear vs normally distributed etc).

As noted in our results, adding covariate information to our Yellowtail Flounder VAST models always decreased the absolute error in the resulting abundance estimates. However, when a reduced number of strata are sampled for Haddock and Cod, adding covariate information leads to a decrease in performance for fall estimates. The decline in performance can be explained by a failure to sample the full spectrum of temperature values where the species exists, leading to an incomplete estimation of the covariate response. Figure \ref{fig:strata-plot} shows the strata excluded from sampling and Figure \ref{fig:CovResponse} depicts the resulting covariate response for Yellowtail Flounder and Haddock for the same population scenario. Although strata were excluded for both species, the survey samples for Yellowtail Flounder still ranged across the species preferred values and thus fully formed the covariate response, while the Haddock covariate response did not contain enough samples from the lower temperature range due to the specific strata that were excluded from sampling (ie, northern strata). As a result, when projecting into those strata as seen in the lower portion of Figure \ref{fig:CovResponse}, Yellowtail estimates with covariate information approximate the correct trend despite these strata being excluded from sampling, while the Haddock estimates that include strata provide that wrong trend and estimate the wrong magnitude of biomass.

@@

{\centering \includegraphics[width=0.9\linewidth]{Images/ratio_trend_YTF} \includegraphics[width=0.9\linewidth]{Images/ratio_trend_cod} \includegraphics[width=0.9\linewidth]{Images/ratio_trend_had}

@@

\caption{Representative example of typical ratio trend for each species, as shown on a log scale.}\label{fig:RatioTrend}
\end{figure}
